---
output: github_document
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  warning=FALSE,
  message = FALSE,
  fig.width=5,
  fig.height=7.5
)
```

# sibregresr


This package is for forecasting salmon returns using sibling regression. The core functions produce model average predictions (i.e., ensembles) of dynamic linear models (DLMs) with sibling predictors. The package also has functionality to forecast using DLMs with penalized-complexity priors, which can be an alternative to model averaging.

See the *Overview* reference article for more details and information on package functionality.

## Installation

You can install sibregresr from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("wdfw-fp/sibregresr")
```

## Example

Below is an example of fitting an ensemble of 7 variations on a sibling regression DLM and evaluating performance over 15 years. A beta version plotting and table function are called to view the ensemble forecasts with weights based on mean absolute percent errors.

See the *Overview* reference article for more details and examples.

```{r example}
library(sibregresr)

forecast<-forecast_fun(
  df = summer_chinook_2024,
  include = c("constIntOnly", "tvIntOnly", "tvSlope", "constLM",
    "tvCRzeroInt", "constCRzeroInt", "tvInt"),
  transformation = log,
  inverse_transformation = exp,
  scale_x = FALSE,
  scale_y = FALSE,
  perf_yrs = 15,
  wt_yrs = NULL,
)

make_table(forecast$forecasts,"MAPE_weight")

make_plot(forecast$forecasts,"MAPE_weight")
```

