---
title: "Overview"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning=FALSE,
  cache = TRUE,
  out.width = "100%",
  cache.extra = 3,
  message = FALSE
  
)
```

## Introduction

This package is meant to forecast salmon returns using sibling regression. The core functions produce model average predictions (i.e., ensembles) of dynamic linear models (DLMs) with sibling predictors. The package also has functionality to forecast using DLMs with penalized-complexity priors, which can be an alternative to model averaging.

## Installation

You can install the development (the only) version of sibregresr from [GitHub](https://github.com/dfw-wa/sibregresr) with:

```{r eval=FALSE}
install.packages("devtools")
devtools::install_github("dfw-wa/sibregresr")
```

```{r setup}
library(sibregresr)
```

## Getting started

### Data format

One must provide a data frame of age-specific returns by either brood- or return year, using specific column names. For example:

```{r}
summer_chinook_2024 |> head(5)
```

or

```{r}
brood_to_return(summer_chinook_2024)|> head(5)
```

If multiple Stocks are included in the data, forecasting will be conducted for each of the stocks. **Returns of the youngest age class are not produced**, but rather are used as a predictor for forecasting the second youngest age class. To forecast the youngest age class (e.g. "Age3"), the data would need to be augmented with a "dummy" column of a yet younger age class (e.g., "Age2").

### Forecasting

Forecasts can be developed using the `forecast_fun()` function.

```{r warning=FALSE}

forecast<-forecast_fun(
  df = summer_chinook_2024,
  include = c("constIntOnly", "tvIntOnly", "tvSlope", "constLM",
    "tvCRzeroInt", "constCRzeroInt", "tvInt"),
  transformation = log,
  inverse_transformation = exp,
  scale_x = FALSE,
  scale_y = FALSE,
  perf_yrs = 15,
  wt_yrs = NULL,
)

```

The arguments in the above call are set to the defaults, which includes 7 models in the ensembles, log transforms the response and predictor, and does not scale the predictor and response prior to fitting. Models can be fit in normal space by specifying `transformation = identity` and `inverse_transformation = identity`. The response and predictors can be Z-scored prior to fitting by specifying `scale_y = TRUE` and `scale_x = TRUE`.

The `perf_yrs` argument specifies how many one-year-ahead forecasts to conduct in order to assess performance of the model. The `wt_yrs` argument specifies how many one-year-ahead forecasts to use for performance-based model averaging (e.g., weighting based on MAPE). When `wt_yrs` is NULL, the number of `perf_yrs` is also used for `wt_yrs`. However, they don't encessarily need to be the same. One could specify `perf_yrs = 15` and `wt_yrs = 1` to conduct a 15-year performance evaluation of the approach of weighting models based on their performance solely in the previous year. It is also possible to use a stretching window rather than a sliding window (i.e., include all available years for performance and/or weighting) by setting `perf_yrs` and/or `wt_yrs` at or above the total number of years in the data.

If `perf_yrs = 15` and `wt_yrs = 15` then `perf_yrs` + `wt_yres` + 1 = 31 years of one-step ahead forecasts for each component (i.e., not the ensemble) model are needed for performance-based model weighting. This is because the first of the 15 years of performance-based ensemble forecasts uses weights based on performance of the components in the previous 15 years. A call to `forecast_fun()` will produce a warning if the number of forecasts needed is greater that the number of data points minus five (assuming at least five years should be available to produce the first forecast).

### Output

The output of a call to the `forecast_fun()` function is a list of two data frames. The "fits" data frame has information on the model fitting that could be useful if some of the models failed to fit. The "forecasts" data frame has a row for every year, age, stock, and model, including model averages. Included in the data frame are the observed and forecasted returns as well as several performance metrics. Definitions for all of the variables in the data frame are prodived in the documentation for the `performance_weights()` function, which can be accessed by a call to `?performance_weights()`. The total forecast for returns summed across ages for a given model are included with the sum of the component ages in the "Age" field. Fore example, when forecasting returns of ages 4, 5, and 6, the total would have a value of 15 in the "Age" field.

Here is an example of some of the information available in the "forecasts" data frame -- the predictions of all the different ensemble models for the most recent year, ordered by MAPE for each year, and some performance metrics.

```{r}
forecast$forecasts |> dplyr::filter(ReturnYear==max(ReturnYear),grepl("weight",model_name)) |> dplyr::select(Stock,Pred,MAPE,RMSE,MPE,MEr) |> dplyr::arrange(Stock,Age,MAPE) |> dplyr::mutate(dplyr::across(where(is.numeric),round))

```

#### Prediction intervals

Prediction intervals are based on the standard deviation between one-year ahead predictions (i.e., forecasts) and observations in log space. The number of years included in the standard deviation calculations is the minimum of the number of years available or the values specified for `perf_yrs`. For example, if `perf_yrs = 15` but only three years of an ensemble forecast had been generated prior to a given year, those three years would be used to calculate the forecast standard deviation and prediction intervals, however, if 20 years of the ensemble forecast were available, only the most recent 15 years would be used. The "n_sd" field in the output of a call to `forecast_fun()` provides the number of years used to calculate prediction intervals for that forecast. The number of years used to calculate weights for an ensemble forecast is conducted in the same way (i.e., all available up to "wt_yrs") and the "n_wts" field in the output gives the number of years used. This is not relevant for ensembles based on AICc weighting as they are based on the AICc value for models fit for the current year.

### Plotting and reporting (beta version)

I've also included a plotting function for predictions from a given model with 50% and 90% prediction intervals. I haven't spent much time refining this though, so it is an area for further development.

```{r fig.width=7,fig.height=9.5,dpi=300}
make_plot(forecast$forecasts,"MAPE_weight")
```

Also a function for a high-level table.

```{r}
make_table(forecast$forecasts,"MAPE_weight")
```

I've started a report template, but quite a bit more work is probably needed. If the forecasting has already been conducted with a call to `forecast_fun()` as above, you can generate the report with a call like:

```{r eval=FALSE}

forecast_report(forecast$forecasts,
                mod_name = "MAPE_weight",
                output_file = "Forcast report.docx"
                )

```

but you can also feed the `forecast_report()` function the raw data and the arguments that you would feed to `forecast_fun()` in order to do it all in one function call. The disadvantage of this is that you will only get the output for one model (e.g., "MAPE_weight") wheras if you fit save the output from a call to `forecast_fun()`, you can investigate them all and render report for multiple models if desired without having to refit.

## Under the `Forecast_fun()` hood

The `Forecast_fun()` is a wrapper around the `mod_funs()`,`setup_data()`,`fit_mods()`, and `performance_weights()` functions.

-   `mod_funs()` generates a list of functions that fit individual models. This list is passed to,
-   `setup_data()`, which generates a nested data frame with a row for each combination of subset of the data (i.e., with some number of years cutoff the end) and model. Each row represents a model that must be fit and prediction made, and this function kicks off the process using a list-column workflow.
-   The `fit_mods()` function does a few data transformation steps and conducts the model fitting.
-   Finally, the `performance_weights()` function uses the fitted model objects to produce forecasts, assesses performance, generates ensembles, and generates preduction intervals.

A similar output as generated by `Forecast_fun()` above could be generated by the following code

```{r eval=FALSE}
mod_list<-mod_funs(c("constIntOnly", "tvIntOnly",
                     "tvSlope", "constLM", "tvCRzeroInt", "constCRzeroInt", "tvInt"))

setup<-setup_data(summer_chinook_2024,mod_list,n_forecasts = 31)

fits<-fit_mods(setup)

forecasts<-performance_weights(fits,perf_yrs = 15)

```

with the only difference being that the `Forecast_fun()` function appends the observations in years for which forecasts were not produced to the output.

## Penalized complexity models

Dynamic linear models with penalized complexity can be fit by including "PenDlm". **It is highly recommended to set `scale_x = TRUE` and `scale_y = TRUE` when fitting the penalized models.** `wt_yrs = 1` below to reduce the number of subsets of the data to fit models to because there is no model averaging happening.

```{r}
pen_dlm_forecast<-forecast_fun(
  df = summer_chinook_2024,
  include = c("PenDlm"),
  transformation = log,
  inverse_transformation = exp,
  scale_x = TRUE,
  scale_y = TRUE,
  perf_yrs = 15,
  wt_yrs = 1,
)

pen_dlm_forecast$forecasts |> dplyr::filter(ReturnYear==max(ReturnYear),model_name=="PenDlm") |> dplyr::select(Stock,Pred,MAPE,RMSE,MPE,MEr) |> dplyr::arrange(Stock,Age,MAPE)|> dplyr::mutate(dplyr::across(where(is.numeric),round))
```

The penalized complexity model puts penalties on the across-year means $\bar{\beta}$ of each coefficient $\beta_t$ for each year $t$, and the standard deviation of the steps in the random walk. So if the coefficients are:

$$
\begin{aligned}
\beta_t &= \beta_{t-1} + \omega_t \\
\omega_t &\sim \mathcal{N}(0,\sigma)
\end{aligned}
$$

$$
\bar{\beta}\sim \mathcal{N}(0,\tau)
$$

This model puts exponential penalties on all $\tau$ and $\sigma$ parameters, for which there is a unique parameter for each predictor in the model:

$$
\begin{aligned}
\sigma,~\tau &\sim \exp(\lambda) \\
\end{aligned}
$$

the default value of $\lambda$ is 1 but can be tuned by the user.

Additionally, $0.01*\text{log}(\tau, \sigma)$ for all $\sigma$s and $\tau$s is added to the log-likelihood to keep their values from shrinking so small as to cause numerical problems during optimization.

### Non-sibling (e.g., climate) predictors

One can change the predictors for the penalized dlm model by providing a data frame of predictor values by return year with a column called "ReturnYear" or by brood year with "BroodYear" for joining with the salmon return data. For example

```{r}

covariates_24 |> head(5)


```

The model handles missing covariate values by fitting them as random effects with a $\mathcal{N}(0,1)$ prior.  

The formula for the model is provided to the `mod_funs()` function and the covariate data to the `fit_mods()` function, but both can be passed through the `forecast_fun()` function as well. When specifying the formula, "x" is used to represent the sibling predictor. So the formula specified below would include the sibling predictor as well as PDO and NPGO from two years prior to when the fish return. In theory, the penalized complexity priors will shrink the coefficients on colinear predictors, so it is ok to include them.

```{r}
pen_dlm_forecast_cov<-forecast_fun(
  df = summer_chinook_2024,
  include = c("PenDlm"),
  transformation = log,
  inverse_transformation = exp,
  scale_x = TRUE,
  scale_y = TRUE,
  perf_yrs = 15,
  wt_yrs = 1,
  covariates = covariates_24,
  penDLM_formula =formula("y~ x + lag2_PDO + lag2_NPGO")
)

pen_dlm_forecast_cov$forecasts |> dplyr::filter(ReturnYear==max(ReturnYear),model_name=="PenDlm") |>
  dplyr::mutate(model_name="PenDlm_cov") |> 
dplyr::select(Stock,Pred,MAPE,RMSE,MPE,MEr) |> dplyr::arrange(Stock,Age,MAPE)|> dplyr::mutate(dplyr::across(where(is.numeric),round))

```

The shape and scale parameters of the gamma prior can be controlled with the `penDLM_gamma_shape` and `penDLM_gamma_scale` arguments to `mod_funs()` or `forecast_fun()`. The fraction of the log of the $\sigma$s that is added to the log-likelihood can be controlled with the `penDLM_regu` argument.

## More general description of the modeling process

The first step in forecasting the returns of a given stock $y_{a,t}$ of age $a$ in year $t$ is to fit models to observed returns through year $t-1$. The models available in this package are variations of a sibling regression model with time-varying intercept and slope:

$$
\left\{
\begin{aligned}
\mathrm{log}(y_{a,t}) & = \alpha_t + \mathrm{log}(y_{a-1,t-1}) \beta_t + v_t, & \quad & v_t \sim \mathcal{N}\left(0, V_t\right), \\
\alpha_t & = \alpha_{t-1} + w_{\alpha, t}, & \quad & w_{\alpha, t} \sim \mathcal{N}\left(0, W_{\alpha, t}\right), \\
\beta_t & = \beta_{t-1} + w_{\beta, t}, & \quad & w_{\beta, t} \sim \mathcal{N}\left(0, W_{\beta, t}\right),
\end{aligned}
\right.
$$

where $\alpha_t$ and $\beta_t$ are an intercept and a slope for the log-transformed returns of the previous age in the previous year, respectively, and both are allowed to vary across years as random walks with process error variances $w_{\alpha, t}$ and $w_{\beta, t}$ respectively. The residual $v_t$ is assumed to be normally distributed around zero with variance $V_t$. Together with a vague prior distribution for the values of $\alpha_0$ and $\beta_0$ (the slope and intercept prior to the first year) these equations define the "full" sibling regression models. However, this "full" model may be overly complex for optimal prediction, and is difficult to fit in practice.

Other models are simplified versions of the full model where:

1)  the slope is assumed to constant through time (i.e., $w_{\beta, t}=0$);\
2)  the intercept is assumed to be constant through time;\
3)  the slope and intercept are assumed to be constant through time;\
4)  the intercept is assumed to be zero (i.e., $\alpha_t=0$);\
5)  the intercept is assumed to be zero and the slope is assumed to be constant through time;\
6)  the slope is assumed to be zero; and\
7)  the slope is assumed to be zero and the intercept is assumed to be constant through time.

These models are fit using the `dlm` package (Petris 2010) within this `sibregresr` package. The penalized version of the full DLM is fit using `RTMB` (Kristensen 2024).

Ensemble forecasts are generated by taking a weighted average of predictions from multiple model. Ensemble model weights are calculated for each model $m$ in the set ($\mathcal{M}$) of 8 models based on their AICc,

$$
w_m=\frac{e^{-0.5\left(\mathrm{AIC}_m-\mathrm{AIC}_{\min }\right)}}{\sum_{i \in \mathcal{M}} e^{-0.5\left(\mathrm{AIC}_i-\mathrm{AIC}_{\min }\right)}},
$$

or based on their observed performance in previous years as measured by the reciprocal of mean absolute percent error (MAPE), root mean square error (RMSE), or mean symmetric accuracy (MeanSA). For example,

$$
w_m=\frac{\mathrm{MAPE}_m^{-1}}{\sum_{i \in \mathcal{M}} \mathrm{MAPE}_i^{-1}}
$$

### References

Giovanni Petris (2010). An R Package for Dynamic Linear Models. Journal of Statistical Software, 36(12), 1-16. URL <https://www.jstatsoft.org/v36/i12/>.

Kristensen K (2024). RTMB: 'R' Bindings for 'TMB'. R package version 1.6, URL <https://github.com/kaskr/RTMB>.

Petris, Petrone, and Campagnoli. Dynamic Linear Models with R. Springer (2009).
